### 2020-06-19 submission/manuscript.Rmd
small edits

To address this need, we noted the similarity between the feeling of unease in analyzing the data and the struggles scientists face with engaging the voluminous scientific literature. A common strategy for keeping up with the literature is by participating in a journal club. Regardless of the discipline, a Journal Club involves group discussion of a pre-selected paper. These range from informal discussions to PowerPoint presentations and course credit. In addition to building upon material from traditional coursework and staying current on the literature, journal clubs help strengthen skills in critical thinking, communication, and integrating the literature [@Lonsdale2016]. Since most journal clubs occur on a regular schedule, they are effective because they incorporate repeated deliberate practices. With this model in mind, over the past four years we have experimented with creating a Code Club model with the goal of improving reproducible data analysis skills in a community environment.

@@

From these experiences, we began a collective conversation to improve our Code Club model. We have identified two successful approaches. The first is a more constructive version of a group code review. The presenter clearly states the problem they want to solve, breaks the participants into smaller groups, and then asks each group to solve the problem, or a portion of it. For example, someone may have an R script with a repeating chunk of code. The challenge for the session would be to convert the code chunk into a function to be called throughout the script to make it "DRY" (i.e. Don't Repeat Yourself [@Wilson2014]). The presenter leaves the session with several partial or working solutions to their problem and the importance of writing DRY code is reinforced. The second approach is a tutorial. The presenter introduces a new package or technique and assigns an activity to practice the new approach. For example, at one Code Club participants were given raw data and a finished plot. Paired participants were tasked with generating the plot from the data using either R syntax from the base language or the ggplot2 package. For this exercise, base R users had to use ggplot2 and vice versa. In either approach, the Code Club ends with a report back to the larger group describing the approach each pair took. We generally find that preparing a Code Club session takes as much effort as preparing to present at a journal club. Our Code Clubs typically have 7 to 10 participants, but the inherent "think-pair-share" approach should allow it to be scaled to larger groups [@Kothiyal2014].

We continue to experiment with approaches for running Code Club. For example, during the period of the Covid-19 pandemic in 2020 when many research labs were shuttered, PDS started posting Code Club materials (https://www.riffomonas.org/code_club/) that individuals or research groups can use to practice their coding skills. During this time, we continued our group's Code Club sessions by making use of the breakout room and screen sharing features in Zoom. Regardless of our approach to leading a Code Club, we have learned that it is critical for the presenter to clearly articulate their goals and facilitate participant engagement. Although some Code Club sessions may be more experimental than others, on the whole they are a critical tool to train bench scientists in reproducible data analysis practices. We have provided some examples of successful topics in Table 1. We have summarized the results of our own experimentation as Code Club presenters and participants into Ten Rules. The first 3 rules apply to all participants, Rules 4 through 8 are targeted to presenters, and the last two for non-presenters. As the rules describe, this model can easily be adapted to groups of non-biologists and to groups with stronger background in computer science.

@@

Our Code Club needs to fit within an hour time slot. When considering their Code Club activity, the presenter should plan for an introduction and brief instruction, time for participants to engage the material, and time for everyone to report back within that hour. A typical schedule for Code Club is 10 minutes of introduction and instruction, 30 minutes of paired programming, 5 minutes to get groups to wrap up, and 10 minutes to report back to the group. We once had a presenter try to teach basic, but unfamiliar, Julia syntax. Unfortunately, the time was up before the participants had installed the interpreter. Some tips to help keep it simple are to limit the presented code chunks to less than 50 lines or, conversely, consider the number of lines that might be required to accomplish a solution. Remember that learners may need up to three times as long to complete a task that is simple for the presenter, so Code Club is best kept simple.

@@

Similar to a Journal Club, the presenter should give participants a few days (ideally a week) to prepare for the Code Club. Considering the compressed schedule described in Rule 5, asking participants to download materials beforehand is helpful to ensure a quick start. The presenter should provide the participants with instructions on how to install dependencies, download data, and get the initial code. This might also uncover weak points in the presenter's plan and enable them to ensure that the materials work as intended before the Code Club. We have found that using GitHub repositories for each Code Club can help make information, scripts, and data easily available to participants. Although this can be convenient, introducing GitHub on top of the session's activities can impose a significant cognitive load and frustration to new group members as well as to the presenter if they are not comfortable with GitHub. Perhaps a first Code Club could introduce using git and GitHub to engage in collaborative coding. A lower barrier entry point for the presenter is posting their code, data, and information in a lab meeting-dedicated Slack channel or via email. Whatever method is used, the presenter should be sure to communicate the topic and necessary materials with the participants ahead of time.

@@

As suggested by Rule 7, a significant challenge to presenting at Code Club is selecting topics and activities that appeal to a critical mass of the participants. This is particularly difficult if participants have a wide range of coding experience, which can follow a turnover in group membership. Beginners will benefit from sessions that cover fundamental concepts and functions. The benefits experienced by more experienced participants include the improved understanding of concepts that comes through teaching the concepts and breaking down problems into simpler elements. Furthermore, instead of focusing on core functions, the group could balance Code Club sessions that cover basics with those that cover methods and packages that are new to everyone in the group. We have also identified several strategies to overcome the challenges presented by participants at varying skill levels. Central to the Code Club format is the use of paired-programming [@Hanks2011]. Instead of letting participants form their own pairs, the presenter can select pairs of participants with either similar or differing skill levels, depending on the presenter's goals. Partnerships between those with similar skill levels requires the presenter to design appropriate activities for each skill level. We have found that commenting code is a good skill for beginners since it forces them to dissect and understand a code chunk line-by-line. It also reinforces the value of commenting as they develop their skills and independence. An advantage of forming partnership between people with disparate skill levels is that it is more likely for groups to provide the presenter with a diverse range of methods that achieve the same result. This approach to pairing also helps to graft new members that have emerging programming skills. Regardless of how partners are selected, consider asking the pairs to identify a navigator and a driver [@Bockeler_2020]. The driver types at the computer while the navigator tells them what to type, thus ensuring participation of both partners. Midway through the activity, the presenter can have the partners switch roles. Intentionally forming pairs can also engineer group interactions by avoiding potentially disruptive partnerships or pairing reliable role models with new group members.

@@

The most important rules are the first and last. Members of the Code Club need to feel comfortable with other group members and sufficiently empowered to try something new or ask for help. Aside from expanding our programming skills, we have noticed two other benefits that help create a positive culture. First, we have intentionally interviewed postdoc candidates on the days we hold Code Club. We make it clear that they are not being assessed on their coding skills, but instead use it as an opportunity to see how the candidate interacts with other members of the research group. At the same time, the candidate can learn about the culture of the research group through active participation. Second, members of other research groups have integrated themselves into our Code Club to minimize the isolation they feel in growing their skills within smaller research groups. This speaks to both the broader need for Code Club and the likelihood of success when expanded to include a larger group of individuals with broader research interests. There is no reason that the Code Club format would not work with a group of people independent of a research group as long as the follow the Rules. Finding datasets and applications that are interesting to a critical mass of people is likely critical to starting and sustaining such a group. Ultimately, Code Club has improved the overall data analysis skills, community, and research success of our lab by empowering researchers to seek help from their colleagues.

>>>>

To address this need, we noted the similarity between the feeling of unease in analyzing the data and the struggles scientists face with engaging the voluminous scientific literature. A common strategy for keeping up with the literature is by participating in a journal club. Regardless of the discipline, a Journal Club involves group discussion of a pre-selected paper that range from informal discussions to PowerPoint presentations and course credit. In addition to building upon material from traditional coursework and staying current on the literature, journal clubs help strengthen skills in critical thinking, communication, and integrating the literature [@Lonsdale2016]. Since most journal clubs occur on a regular schedule, they are effective by virtue of repeated practice. With this model in mind, over the past four years we have experimented with creating a Code Club model with the goal of improving reproducible data analysis skills in a community environment.

@@

From these experiences, we began a collective conversation to improve our Code Club model. We have identified two successful approaches. The first is a more constructive version of a group code review. The presenter clearly states the problem they want to solve, breaks the participants into smaller groups, and then asks each group to solve the problem, or a portion of it. For example, someone may have an R script with a repeating chunk of code. The challenge for the session would be to convert the code chunk into a function to be called throughout the script to make it "DRY" (i.e. Don't Repeat Yourself [@Wilson2014]). The presenter leaves the session with several partial or working solutions to their problem and the importance of writing DRY code is reinforced. The second approach is a tutorial. The presenter introduces a new package or technique and assigns an activity to practice the new approach. For example, at one Code Club participants were given raw data and a finished plot. Paired participants were tasked with generating the plot from the data using R syntax from either the base language or the ggplot2 package. For this exercise, base R users had to use ggplot2 and vice versa. In either approach, the Code Club ends with a report back to the larger group describing the approach each pair took. We generally find that preparing a Code Club session takes similar effort to preparing a journal club presentation. Our Code Clubs typically have 7 to 10 participants, but the inherent "think-pair-share" approach should allow it to be scaled to larger groups [@Kothiyal2014].

We continue to experiment with approaches for running Code Club. For example, during the early period of the Covid-19 pandemic in 2020 when many research labs were shuttered, PDS started posting Code Club materials (https://www.riffomonas.org/code_club/) that individuals or research groups can use to practice their coding skills. During this time, we continued our group's Code Club sessions using the breakout room and screen sharing features in Zoom. Regardless of our approach to leading a Code Club, we have learned that it is critical for the presenter to clearly articulate their goals and facilitate participant engagement. Although some Code Club sessions may be more experimental than others, on the whole they are a critical tool to train bench scientists in reproducible data analysis practices. We have provided some examples of successful topics in Table 1. We have summarized the results of our own experimentation as Code Club presenters and participants into Ten Rules. The first 3 rules apply to all participants, Rules 4 through 8 are targeted to presenters, and the last two for non-presenters. As the rules describe, this model can easily be adapted to groups of non-biologists or those with stronger backgrounds in computer science.

@@

Our Code Club needs to fit within an hour time slot. When considering their Code Club activity, the presenter should plan for an introduction and brief instruction, time for participants to engage the material, and time for everyone to report back within that hour. A typical schedule for Code Club is 10 minutes of introduction and instruction, 30 minutes of paired programming, 5 minutes to get groups to wrap up, and 10 minutes to report back to the group. We once had a presenter try to teach basic, but unfamiliar, Julia syntax. Unfortunately, the time was up before the participants had installed the interpreter. Some tips to help keep it simple are to limit the presented code chunks to less than 50 lines or, conversely, consider the number of lines that might be required to accomplish a solution. Remember that learners may need up to three times as long to complete a task that is straight forward for the presenter, so Code Club is best kept simple.

@@

Similar to a Journal Club, the presenter should give participants a few days (ideally a week) to prepare for the Code Club. Considering the compressed schedule described in Rule 5, asking participants to download materials beforehand is helpful to ensure a quick start. The presenter should provide the participants with instructions on how to install dependencies, download data, and get the initial code. This might also uncover weak points in the presenter's plan and enable them to ensure that the materials work as intended before the Code Club. We have found that using GitHub repositories for each Code Club can help make information, scripts, and data easily available to participants. Although this can be convenient, introducing GitHub on top of the session's activities can impose a significant cognitive load and frustration to those not already comfortable with GitHub. Perhaps a first Code Club could introduce using git and GitHub to engage in collaborative coding. A lower barrier entry point for the presenter is posting their code, data, and information in a lab meeting-dedicated Slack channel or via email. Whatever method is used, the presenter should be sure to communicate the topic and necessary materials with the participants ahead of time.

@@

As suggested by Rule 7, a significant challenge to presenting at Code Club is selecting topics and activities that appeal to a critical mass of the participants. This is particularly difficult if participants have a wide range of coding experience, which can follow a turnover in group membership. Beginners will benefit from sessions that cover fundamental concepts and functions. The benefits for more experienced participants include the improved understanding of concepts by teaching and breaking down problems into simpler elements. Furthermore, instead of focusing on core functions, the group could balance Code Club sessions that cover basics with those that introduce new methods and packages to the group. We have also identified several strategies to overcome the challenges presented by participants at varying skill levels. Central to the Code Club format is the use of paired-programming [@Hanks2011]. Instead of letting participants form their own pairs, the presenter can select pairs of participants with either similar or differing skill levels, depending on the presenter's goals. Partnerships between those with similar skill levels requires the presenter to design appropriate activities for each skill level. We have found that commenting code is a good skill for beginners since it forces them to dissect and understand a code chunk line-by-line. It also reinforces the value of commenting as they develop their skills and independence. An advantage of forming partnership between people with disparate skill levels is that it is more likely for groups to provide the presenter with a diverse range of methods that achieve the same result. This approach to pairing also helps to graft new members that have emerging programming skills. Regardless of how partners are selected, consider asking the pairs to identify a navigator and a driver [@Bockeler_2020]. The driver types at the computer while the navigator tells them what to type, thus ensuring participation of both partners. Midway through the activity, the presenter can have the partners switch roles. Intentionally forming pairs can also engineer group interactions by avoiding potentially disruptive partnerships or pairing reliable role models with new group members.

@@

The most important rules are the first and last. Members of the Code Club need to feel comfortable with other group members and sufficiently empowered to try something new or ask for help. Aside from expanding our programming skills, we have noticed two other benefits that help create a positive culture. First, we have intentionally interviewed postdoc candidates on the days we hold Code Club. We make it clear that they are not being assessed on their coding skills, but instead use it as an opportunity to see how the candidate interacts with other members of the research group. At the same time, the candidate can learn about the culture of the research group through active participation. Second, members of other research groups have integrated themselves into our Code Club to minimize the isolation they feel in growing their skills within smaller research groups. This speaks to both the broader need for Code Club and the likelihood of success when expanded to include a larger group of individuals with broader research interests. There is no reason that the Code Club format would not work with groups independent of a research group as long as they follow the Rules. Finding datasets and applications that are interesting to a critical mass of people will likely be key to starting and sustaining such a group. Ultimately, Code Club has improved the overall data analysis skills, community, and research success of our lab by empowering researchers to seek help from their colleagues.
